---
title: ""
output:
  html_document: default
  pdf_document: default
date: "2025-04-19"
---

```{r setup, include=FALSE}

# Install packages
library(pacman)
p_load(tinytex, tidyverse,readxl,janitor,stringr,RColorBrewer)
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 6, tidy = TRUE)

# Aggregate to county level; 
counties_of_interest <- c("Los Angeles County", "Orange County", "Sacramento County", "San Diego County", "San Francisco County", "Santa Cruz County")

# since we are working over a study period of five years, it is best to automate the process of importing the data

# master datasets: 
datasets <- c(
  "ACS Data/Web_ACS2018_Housing.xlsx",
  "ACS Data/Web_ACS2019_Housing.xlsx",
  "ACS Data/Web_ACS2021_Housing.xlsx",
  "ACS Data/Web_ACS2022_Housing.xlsx",
  "ACS Data/Web_ACS2023_Housing.xlsx"
)

# TENURE SPREADSHEET
# Description: Gives us the number and percentage of renters with margin of error
tenure_list <- list()

for (path in datasets) {
  year <- stringr::str_extract(path, "\\d{4}")
  df <- read_excel(path, sheet = "Tenure", skip = 4) %>%
    clean_names() %>%
    filter(geography %in% counties_of_interest) %>%
    mutate(year = year,
           lower_nCI = renter_occupied_estimate - renter_occupied_margin_of_error,
           upper_nCI = renter_occupied_estimate + renter_occupied_margin_of_error,
           lower_pCI = renter_occupied_percent - renter_occupied_percent_margin_of_error,
           upper_pCI = renter_occupied_percent + renter_occupied_percent_margin_of_error)
  tenure_list[[year]] <- df
}

tenure_all <- bind_rows(tenure_list)

# RENTER COSTS SPREADSHEET
# Description: Gives us the GROSS RENT and GRAPI Data 
rentercost_list <- list()

for (path in datasets) {
  year <- stringr::str_extract(path, "\\d{4}")
  df <- read_excel(path, sheet = "Renter Costs", skip = 4) %>%
    clean_names() %>%
    filter(geography %in% counties_of_interest) %>%
    mutate(year = year,
           across(2:37, ~ as.numeric(.)),
           gross_rent_lower = gross_rent_median_dollars_estimate - gross_rent_median_dollars_margin_of_error,
           gross_rent_upper = gross_rent_median_dollars_estimate + gross_rent_median_dollars_margin_of_error)
  rentercost_list[[year]] <- df
}

rentercosts_all <- bind_rows(rentercost_list)

# Sacramento
tenure_sac <- tenure_all %>% 
  filter(geography == "Sacramento County")

renterscost_sac <- rentercosts_all %>% 
  filter(geography == "Sacramento County")

```


$Costs^{1,2}$

Eviction timeline: 90-120 days (factor into lost rent for the period)

Court filing costs: \$240-\$450

Writ of Possession: \$25-\$40

Sheriff Fees: \$100-\$150

Process Server Fees: \$50-200

Attorney Fees: \$500-\$2000 +

Property Damage: case specific

Storage / Disposal of Property: case specific

**Total estimated minimum cost: $3000 - $4000**


```{r nRent, echo=FALSE}
Nrenters <- tenure_sac$renter_occupied_estimate

ggplot(data = tenure_sac, aes(x = year, y = Nrenters)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_errorbar(aes(ymin = lower_nCI, ymax = upper_nCI), width = 0.2) + 
  labs(title = "Renting Households over Time in Sacramento",
       x = "Year",
       y = "Number of Renting Households",
       caption = "Source: ACS Data") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 18),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")

```


```{r pRent, echo=FALSE}
Prenters <- tenure_sac$renter_occupied_percent

ggplot(data = tenure_sac, aes(x = year, y = Prenters)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_errorbar(aes(ymin = lower_pCI, ymax = upper_pCI), width = 0.2) + 
  labs(title = "Percentage of Renting Households over Time in Sacramento",
       x = "Year",
       y = "Percentage of Renting Households",
       caption = "Source: ACS Data") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 18),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")
```

_____________________________________________________________________________________________________________________
#### Analysis: 
A quick glance at these may suggest a decrease in risk to the property management companies due to a lower volume of tenants. On the contrary, we believe that this decline in both percentage and number of renting households indicates economic hardship. It is possible that this decrease in volume is driven by a consolidation in rentals. Such examples of this include people moving back to their parents' home, moving in with their partner, or increasing the number of lessors in an individual lease. 
_____________________________________________________________________________________________________________________


```{r MedianRent, echo=FALSE}
cpi_multipliers <- c(
  "2018" = 0.9498,
  "2019" = 0.9667,
  #"2020" = 0.9837,
  "2021" = 1.0000,
  "2022" = 1.0395,
  "2023" = 1.0757
)

renterscost_sac <- renterscost_sac %>% 
  mutate(adjusted_median_rent = gross_rent_median_dollars_estimate / cpi_multipliers[as.character(year)],
         adjusted_median_rent_moe = gross_rent_median_dollars_margin_of_error / cpi_multipliers[as.character(year)],
         adj_lower = adjusted_median_rent - adjusted_median_rent_moe,
         adj_higher = adjusted_median_rent + adjusted_median_rent_moe)

# Rent Plot
ggplot(data = renterscost_sac, aes(x = year, y = adjusted_median_rent)) +
  geom_point(size = 2, alpha = 0.7) +
  geom_errorbar(aes(ymin = adj_lower, ymax = adj_higher), width = 0.2) + 
  labs(title = "Price Adjusted Median Rent Price in Sacramento County",
       x = "Year",
       y = "Gross Median Rent ($)",
       caption = "Source: ACS Data") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 18),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")
```

_____________________________________________________________________________________________________________________
#### Analysis: 
Using the Chained CPI index to adjust prices to 2021 dollars, we show that there is a real increase in the median rent price for Sacramento County. Over the period 2018-2023, the median rent price increased 35% (from \$1300 to \$1762). Having controlled for inflation, it is clear that rent is a driving factor in the overall increase in cost of living for Sacramento County. Rentals are becoming less affordable to tenants, thereby increasing property managers' financial risk. 
_____________________________________________________________________________________________________________________



```{r GRAPHI, echo=FALSE}
graphi_columns <- c(
  "gross_rent_as_a_percentage_of_household_income_grapi_less_than_15_0_percent_estimate",
  "gross_rent_as_a_percentage_of_household_income_grapi_15_0_to_19_9_percent_estimate",
  "gross_rent_as_a_percentage_of_household_income_grapi_20_0_to_24_9_percent_estimate",
  "gross_rent_as_a_percentage_of_household_income_grapi_25_0_to_29_9_percent_estimate",
  "gross_rent_as_a_percentage_of_household_income_grapi_30_0_to_34_9_percent_estimate",
  "gross_rent_as_a_percentage_of_household_income_grapi_35_0_percent_or_more_estimate",
  "gross_rent_as_a_percentage_of_household_income_grapi_not_computed_estimate"
)

# Rename Columns
renterscost_sac <- renterscost_sac %>% 
  rename(
    GRAPHI_15 = gross_rent_as_a_percentage_of_household_income_grapi_less_than_15_0_percent_estimate,
    GRAPHI_15_19.9 = gross_rent_as_a_percentage_of_household_income_grapi_15_0_to_19_9_percent_estimate,
    GRAPHI_20_24.9 = gross_rent_as_a_percentage_of_household_income_grapi_20_0_to_24_9_percent_estimate,
    GRAPHI_25_29.9 = gross_rent_as_a_percentage_of_household_income_grapi_25_0_to_29_9_percent_estimate,
    GRAPHI_30_34.9 = gross_rent_as_a_percentage_of_household_income_grapi_30_0_to_34_9_percent_estimate,
    GRAPHI_35 = gross_rent_as_a_percentage_of_household_income_grapi_35_0_percent_or_more_estimate,
    GRAPHI_NC = gross_rent_as_a_percentage_of_household_income_grapi_not_computed_estimate
  )

# Because the data is in long format, we can't readily plot this in ggplot. 
renterscost_sac_graphi <- renterscost_sac %>% 
  pivot_longer(cols = starts_with("GRAPHI_"),
               names_to = "GRAPHI_Category",
               values_to = "GRAPHI_Value")
  

# GRAPHI, Final Plot
ggplot(data = renterscost_sac_graphi, aes(x = year, y = GRAPHI_Value, fill = GRAPHI_Category)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Blues") +
  labs(title = "Gross Rent as a Percentage of Household Income by Year",
       subtitle = "Sacramento",
       x = "Year",
       y = "Gross Rent as a Percentage of Household Income",
       caption = "Source: https://data-downloads.evictionlab.org/#data-for-analysis/") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 18),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right")
```

_____________________________________________________________________________________________________________________
#### Analysis
Gross rent as a percentage of household income. This measure is self explanatory; it is the rent that tenants pay, as a percentage of the household income. 
This figure illustrates the metric in bins, where GRAPHI_15 gives the households whose rent is less than, or equal to 15% of their household income, GRAPHI_15_19.9 is those whose rent is in the range 15% - 19.9%, and so on. A majority of households in this survey throughout the years have rent greater than, or equal to 35% of their income. These are the households who are likely living paycheck to paycheck, and are perhaps one or two bad months away from not being able to make rent in a given month. 
Our analytical methods aim to examine these types of households, and their spending habits closely to make a better determination of whether or not they would constitute a high risk of missing rent. 
_____________________________________________________________________________________________________________________

Source list: 

1. https://www.doorloop.com/laws/california-eviction-process

2. https://www.goodlifemgmt.com/blog/california-eviction-process-cost-and-steps/

3. (ACS Data) https://dof.ca.gov/reports/demographic-reports/american-community-survey/#ACS2018x1



